---
title: Tables
format: gfm
---

This will be a very quick lab where you just practice formatting your data and making a table of it. Not much that isn't already in the notes.

- Use the ACS data, filtered for the US, Maryland, and a few counties of your choice. 
- Choose a few columns you're interested in, and select those (plus location). 
- Use more than one type of numeric variable (choose a mix of percentages, integers, and/or dollar amounts).
- Define the formatter functions you'll need, and use them to format each variable as appropriate.
- Create a table using `knitr::kable`.
- Give your columns clean, readable names.
- Align the columns so text columns are aligned on the left, numeric ones on the right.
- Give the table a title using the `caption` argument of `kable`.


```{r}
#| label: setup

library(ggplot2)

acs_table <- justviz::acs |>
    dplyr::filter(
        name %in%
            c(
                "United States",
                "Maryland",
                "Washington County",
                "Baltimore County",
                "Frederick County",
                "Garrett County"
            )
    ) |>
    dplyr::select(level, name, foreign_born, median_hh_income, diversity_idx, bachelors)

acs_table
 
```


```{r}
#| label: table-with-kable

comma <- scales::label_comma(accuracy = 1) # round to nearest whole number
percent <- scales::label_percent(accuracy = 1)
dollar <- scales::label_currency(accuracy = 1)

acs_table_fmt <- acs_table |>
    dplyr::mutate(foreign_born = percent(foreign_born)) |>
    dplyr::mutate(bachelors = percent(bachelors)) |>
    dplyr::mutate(median_hh_income = dollar(median_hh_income)) |>
    dplyr::mutate(diversity_idx = percent(diversity_idx))
 
knitr::kable(
    acs_table_fmt,
    align = "llrrrr",
    col.names = c(
        "Level",
        "Name",
        "Foreign born",
        "Median household income",
        "Diversity index",
        "Bachelors deg"
    ),
    caption = "Socioeconomic indicators, western MD counties, 2024"
)

```